%%%%%%%%%%%%%%%%%%%%%%%%
%% Sample use of the infthesis class to prepare a thesis. This can be used as 
%% a template to produce your own thesis.
%%
%% The title, abstract and so on are taken from Martin Reddy's csthesis class
%% documentation.
%%
%% MEF, October 2002
%%%%%%%%%%%%%%%%%%%%%%%%

%%%%
%% Load the class. Put any options that you want here (see the documentation
%% for the list of options). The following are samples for each type of
%% thesis:
%%
%% Note: you can also specify any of the following options:
%%  logo: put a University of Edinburgh logo onto the title page
%%  frontabs: put the abstract onto the title page
%%  deptreport: produce a title page that fits into a Computer Science
%%      departmental cover [not sure if this actually works]
%%  singlespacing, fullspacing, doublespacing: choose line spacing
%%  oneside, twoside: specify a one-sided or two-sided thesis
%%  10pt, 11pt, 12pt: choose a font size
%%  centrechapter, leftchapter, rightchapter: alignment of chapter headings
%%  sansheadings, normalheadings: headings and captions in sans-serif
%%      (default) or in the same font as the rest of the thesis
%%  [no]listsintoc: put list of figures/tables in table of contents (default:
%%      not)
%%  romanprepages, plainprepages: number the preliminary pages with Roman
%%      numerals (default) or consecutively with the rest of the thesis
%%  parskip: don't indent paragraphs, put a blank line between instead
%%  abbrevs: define a list of useful abbreviations (see documentation)
%%  draft: produce a single-spaced, double-sided thesis with narrow margins
%%
%% For a PhD thesis -- you must also specify a research institute:
\documentclass[phd,icsa,logo,twoside]{infthesis}

%% For an MPhil thesis -- also needs an institute
% \documentclass[mphil,ianc]{infthesis}

%% MSc by Research, which also needs an institute
% \documentclass[mscres,irr]{infthesis}

%% Taught MSc -- specify a particular degree instead. If none is specified,
%% "MSc in Informatics" is used.
% \documentclass[msc,cogsci]{infthesis}
% \documentclass[msc]{infthesis}  % for the MSc in Informatics

%% Master of Informatics (5 year degree)
% \documentclass[minf]{infthesis}

%% Undergraduate project -- specify the degree course and project type
%% separately
% \documentclass[bsc]{infthesis}
% \course{Artificial Intelligence and Psychology}
% \project{Fourth Year Project Report}

%% Logo Style: University of Edinburgh Shield
% \documentclass[minf]{infthesis}
%% Possible values for shieldtype:
%% 0: regular monochrome
%% 1: monochrome with no background lines
%% 2: reverse monochrome
%% 3: two colours: navy and red
%% 4: full colour
\shieldtype{4}


%% Put any \usepackage commands you want to use right here; the following is 
%% an example:
\usepackage[numbers,square,sort]{natbib}
%\usepackage{bibentry}
%\usepackage{biblatex}

\usepackage{graphicx}
\usepackage{subcaption}

\usepackage{amsmath}
\usepackage{amssymb}


\usepackage{multirow}           %tables
\usepackage{rotating}           %tables

\usepackage{colortbl}
\usepackage{color}
\usepackage{booktabs}
%\usepackage[table]{xcolor}
\definecolor{oddcolor}{gray}{1.0}
\definecolor{evencolor}{gray}{0.9}

%\newcommand{\fixme}[1]{{\color{red}{#1}}}
\newcommand{\fixme}[1]{#1}

\newcommand{\itercomp}{{iterative compilation}}
\newcommand{\Itercomp}{{Iterative compilation}}
\newcommand{\IterComp}{{Iterative Compilation}}

\newtheorem{definition}{Definition}

% \newcommand{\etal}{et~al.}

% \usepackage{minted}
% \usemintedstyle{trac} %vs,borland,pastie,perldoc,friendly

%% Information about the title, etc.
\title{Reducing Code Size with Function Merging}
\author{Rodrigo Caeteano de Oliveira Rocha}

%% If the year of submission is not the current year, uncomment this line and 
%% specify it here:
% \submityear{1785}

%% Optionally, specify the graduation month and year:
% \graduationdate{February 1786}

%% Specify the abstract here.
    
%\include{src/abstract.tex}
\abstract{
%\include{src/abstract}    

Resource-constrained devices for embedded systems are becoming increasingly important.
In such systems, primary and secondary memories are highly restrictive, making code size in most cases
even more important than performance.
Compared to more traditional platforms, memory is a larger part of the cost and
code occupies much of it. Despite that, compilers make little effort to reduce
code size.
One important optimisation for  code-size reduction is function merging.
%One key technique attempts to merge the bodies of similar functions.
This technique eliminates redundant code across functions by
merging them into a single function.

However, production compilers only apply this optimisation to identical functions,
while research compilers improve on that by merging the few functions with identical control-flow graphs and signatures.
Overall, existing solutions are insufficient and we end up having to either increase cost by adding more memory or remove functionality from programs.

This thesis introduces the first techniques capable of merging arbitrary pair of functions.
Our insight is that the weak results of prior function merging techniques
are due to their rigid and overly restrictive solutions instead of the lack of duplicate code in the input program.
Our solution corroborates this insight, resulting in drastic code size reductions while also reducing end-to-end compilation time.

First, we introduce FMSA, a novel technique that can merge arbitrary functions through sequence
alignment, a bioinformatics algorithm for identifying regions of similarity
between sequences. We combine this technique with an intelligent exploration
mechanism to direct the search towards the most promising function pairs.
Our approach is more than 2.4$\times$ better than the previous state of the art, proposed by von Koch~et~al., reducing code size by up to 25\%, with an overall average of 6\%.
FMSA increases end-to-end compilation time by an average of 15\%.

While representing a leap forward, experiments show that FMSA fails to reduce
code size in some cases where it would be intuitively expected to work.
This limitation stems from its inability to directly handle phi-nodes.
Instead, FMSA applies register demotion to replace all such nodes with memory
operations, in an attempt to simplify the code generation
processes.
We build on this technique and develop SalSSA that fully supports the SSA form, removing any need for register demotion.
By doing so, we notably increase the number of profitably merged functions. 
Experimental results on the SPEC 2006 and 2017 suites show that our approach delivers on average, 7.9\% to 9.7\% reduction on the final size of the compiled code.
%This translates to around 2$\times$ more code size reduction over our first solution, FMSA.
Moreover, as a result of aligning shorter sequences of instructions and reducing the number of wasteful merge operations, our new approach incurs an average compile-time overhead of only 5\%,
3$\times$ less than FMSA, while also reducing memory usage by over $2\times$.

Finally, we continue to build on SalSSA by developing HyFM, which delivers similar levels of code size reduction for significantly lower compilation time and memory usage.
To this end, we introduce an alignment strategy that works at the basic block level.
Since basic blocks are usually much shorter than functions, even a quadratic alignment is acceptable. However, we also propose a linear algorithm for aligning blocks at a much lower cost. We extend this strategy with a multi-tier profitability analysis that bails out early from unprofitable merging attempts. By aligning individual pairs of blocks, we are able to decide their alignment's profitability 
before actually generating code.
Experimental results on SPEC~2006 and 2017 show that HyFM needs orders of magnitude less memory, using up to 48~MB or 5.6~MB, depending on the variant used, while SalSSA requires 32~GB in the worst case.
HyFM also runs over 4.5$\times$ faster, while still achieving comparable code size reduction. Combined with the speedup of later compilation stages due to the reduced number of functions, HyFM contributes to a reduced end-to-end compilation time.

}

%% Now we start with the actual document.
\begin{document}

%% First, the preliminary pages
\begin{preliminary}

%% This creates the title page
\maketitle

%% Acknowledgements
\begin{acknowledgements}
\textbf{TODO.}
\end{acknowledgements}

%% Next we need to have the declaration.
%\standarddeclaration
\papersdeclaration{
\begin{itemize}

%\item \fullcite{rocha19} 
%\item \bibentry{rocha19}
%\item \citefullauthor{rocha19}, \citeyear{rocha19}.
\item Rodrigo Rocha, Pavlos Petoumenos, Zheng Wang, Murray Cole, and Hugh Leather. ``Function merging by sequence alignment.'' In IEEE/ACM International Symposium on Code Generation and Optimization (CGO), pp. 149-163. Best Paper Award. 2019.

\item Rodrigo Rocha, Pavlos Petoumenos, Zheng Wang, Murray Cole, and Hugh Leather. ``Effective function merging in the SSA form.'' In ACM SIGPLAN Conference on Programming Language Design and Implementation (PLDI), pp. 854-868. 2020.

\item Rodrigo Rocha, Pavlos Petoumenos, Zheng Wang, Murray Cole, Hugh Leather, Kim Hazelwood. ``HyFM: Function Merging for Free.'' In the International Conference on Languages Compilers, Tools and Theory of Embedded Systems (LCTES). 2021.

\end{itemize}
}

%% Finally, a dedication (this is optional -- uncomment the following line if
%% you want one).
% \dedication{To my mummy.}

%% Create the table of contents
\tableofcontents

%% If you want a list of figures or tables, uncomment the appropriate line(s)
 \listoffigures
 \listoftables

\end{preliminary}

%%%%%%%%
%% Include your chapter files here. See the sample chapter file for the basic
%% format.


\newcommand{\ProjName}{{SalSSA}\xspace} %Sequence ALignment in the SSA form
\newcommand{\etal}{{et~al.}}

\include{src/intro}
\include{src/background}
\include{src/relatedwork}
%\include{src/merge-operation}
\include{src/cgo19}
%\include{src/merging-optimisation}
\include{src/pldi20}
%\include{src/runtime-overhead}
%\include{src/deeplearning}
%\include{src/fastfm}
\include{src/lctes21}
\include{src/conclusion}

%%%%%%%%
%% Any appendices should go here. The appendix files should look just like the
%% chapter files.
%\appendix
%\include{appendix1}
%% ... etc...

%% Choose your favourite bibliography style here.
%\bibliographystyle{apalike}
%\bibliographystyle{plainnat}
\bibliographystyle{plain}

%% If you want the bibliography single-spaced (which is allowed), uncomment
%% the next line.
% \singlespace

%% Specify the bibliography file. Default is thesis.bib.
\bibliography{bibliography.bib}

%% ... that's all, folks!
\end{document}
