
\section{Merging Beyond Identical Functions}

The state-of-the-art function-merging technique exploits structural similarity among functions~\cite{edler14}.
Their optimization is able to merge similar functions that are not necessarily
identical.
Two functions are structurally similar if both their function types are equivalent
and their CFGs isomorphic.
Two function types are equivalent if they agree in the number, order, and types
of their parameters as well as
their return types, linkage type, and other compiler-specific properties.
In addition to the structural similarity of the functions, their technique also
requires that corresponding basic blocks have exactly the same number of instructions
and that corresponding instructions must have equivalent resulting types.
% but may differ in their opcodes or in the number and type of their input operands.
Mergeable functions are only allowed to differ in corresponding instructions,
where they can differ in their opcodes or in the number and type of their input operands.
%The only differences that are actually allowed is that
%corresponding instructions can 
%differ in their opcodes or in the number and type of their input operands.

%If two corresponding instructions have different opcodes, they split the basic
%block and insert a switch branch to select which instruction to execute
%depending on a function identifier.

Because the state-of-the-art is limited to functions with identical CFGs
and function types, once it merges a pair of functions, a third
\textit{similar} function cannot be merged into the resulting merged function
since they will differ in both CFGs and their lists of parameters.
Due to this limiting factor, the state-of-the-art has to first collect all
mergeable functions and merge them simultaneously.

The state-of-the-art algorithm iterates simultaneously over corresponding basic
blocks in the set functions being merged, as they have isomorphic CFGs.
For every basic block, if their corresponding instructions have different opcodes,
they split the basic block and insert a switch branch to select which instruction
to execute depending on a function identifier.
Because these instructions have equivalent resulting types, their results can be
merged using a phi-operator, which can then be used transparently as operands
by other instructions.

Although the state-of-the-art technique improves over LLVM's identical function merging, it is
still unnecessarily limited. In Section~\ref{sec:motivation}, we showed examples of very similar
real functions where the state-of-the-art fails to merge. Our approach addresses such limitations
improving on the state-of-the-art across the board.


