
\chapter{Effective Function Merging in the SSA Form} \label{chp:pldi20}


%Prior function merging methods were limited to identical or isomorphic functions,
%but a recent work has generalized function merging to any arbitrary pair of functions.
%The approach proposed in Chapter~\ref{chp:cgo19} first represents functions as nothing more
%than linear sequences of instructions and labels.
%Then it applies a sequence alignment algorithm, developed for bioinformatics, to discover
%the optimal way to create pairs of mergeable instructions from the two input
%sequences.
%Finally, it generates the merged function where aligned pairs of matching
%instructions are merged to a single instruction, while non-matching instructions
%are simply copied into the merged function.

While the technique proposed in Chapter~\ref{chp:cgo19} represents a leap forward, experiments show that FMSA fails to reduce code size in some cases where it would be intuitively expected to work.
Even when handling similar functions that should be profitably merged, this algorithm may fail spectacularly, producing a merged function \textit{larger} than the combined input functions.
%Even when handling similar functions, this algorithm may fail spectacularly,
%producing a merged function which is \textit{larger} than the combined input
%functions.
%Their algorithm handles this problem by ignoring the merging output when it is
%not profitable, resulting in missed opportunities to reduce code, which should
%not happen.

Closer inspection reveals that the problem stems from the inability of this
approach to handle \textit{phi-nodes}. In SSA, \textit{phi-nodes} merge the
assignments of a single variable that arrive from different control flow paths.
As such, they are closely tied to how control and data flow across basic blocks
and cannot be merged without examining their control flow context.
FMSA generates code directly from the aligned sequences, where
control flow information has been lost, merging instructions blindly with little
to no consideration for their context, so it cannot handle \textit{phi-nodes}.
It overcomes this hurdle by applying register demotion, which replaces
\textit{phi-nodes} with stack variables.
This works but only by artificially increasing the size of the input functions,
often by twice or more their original size, the exact opposite of what function merging
tries to achieve.
A final post-merging step of register promotion is supposed to reverse this code
bloating but it often fails, leading to unprofitable merged functions.

Our idea is to keep the one thing that works well in FMSA, the idea of using sequence alignment on functions, and build
around it a new function merging methodology that can handle directly control and data flow with no need for register demotion. Our proposed approach,
{\ProjName}, achieves this with a new code generator for aligned functions. Instead of translating the alignment directly into a merged function,
our approach generates code from the input control-flow graphs, using the alignment only to specify pairs of matching labels and instructions.
The generator then produces code top-down, starting with the control flow graph of the merged function, then populating with instructions, arguments and labels, and
finally with \textit{phi-nodes} which maintain the correct flow of data.
{\ProjName} is carefully designed to produce correct but, still, succinct code. A final post-generator stage applies a novel optimization,
\textit{phi-node coalescing}, that eliminates superfluous phi-nodes and select instructions, reducing even further the code size.

%This optimization combines disjoint value definitions from both input functions,
%reducing the number of phi-nodes and selection instructions required for the final merged function.
%\textit{Phi-node coalescing} has the benefit of reducing register pressure therefore
%resulting in smaller code.

%\fixme{Mention results about memory usage.}
{\ProjName} produces functions much smaller than those produced by FMSA. In many cases, it produces profitable merged functions where FMSA
fails. On average, it reduces about twice as much code as their approach, 11.4\% to 14.5\% compared to 5.6\% to 6.2\% depending on the
function merging configuration. On top of that, the compile-time overhead is much lower. Sequence alignment has a quadratic relationship
with function size, while the overhead of code generation and later optimization passes is proportional to function size. By avoiding
register demotion, we keep input function sequences smaller and we produce smaller functions, leading to an average compilation overhead of
5\%, $3\times$ less than FMSA, and an overhead in no case more than 55\%, compared to the maximum overhead of 314\% for FMSA. Similarly,
{\ProjName} uses half the amount of memory required on average by FMSA during compilation.

With this chapter, we make the following contributions:
\begin{itemize}
  \item The first approach that fully supports the SSA form when merging functions through sequence alignment.
  %\item A new flexible merged function generator that handles control flow and data flow efficiently.
  \item A novel optimization called phi-node coalescing that reduces the number of phi-nodes and selections in merged functions.
  \item {\ProjName} achieves about twice as much code size reduction than the state of the art with significantly lower compilation time
        overheads.
  %\item {\ProjName} achieves about $2\times$ better function merging for $3\times$ less overhead than the state-of-the-art, making
  %    function merging worthwhile for large code bases.
\end{itemize}


\input{src/pldi20/motivation}

\input{src/pldi20/contribution}
\input{src/pldi20/results}
\input{src/pldi20/conclusion}

